\section{Introduction}
Plotkin and Power's algebraic effects \cite{Plotkin2001} combined with Plotkin and Pretnar's handlers \cite{Plotkin2013} yield a programming model for controlling computational effects. An algebraic effect is a collection of abstract operations without predefined semantics. Handlers interpret algebraic effects by assigning semantics to abstract operations. Thus, the control of computational effects is lifted from the run-time system into the hands of the programmer. Clearly, it shifts additional responsibility onto the programmer, however, in addition it opens up a design space that was previously secluded from the programmer. Now, the programmer may extend the capabilities of the run-time system beyond what the language implementor envisaged.

For example, it is well-known that concurrency primitives can be encoded through algebraic effects \cite{Bauer2015,Dolan2015} as spawning a new thread is an effect. Furthermore, suspending a thread is an effect. Handlers of these effects are schedulers. Hence, the programmer is free to implement a particular scheduling policy tailored for the application. The choice of scheduler can have significant impact on performance as a scheduler's performance goal can be paramount to the overall performance of an application \cite{Berman2003}.

However\dots

\section{Problem definition}

\section{Background}
\subsection{Handlers implementations}
\subsubsection{First-class implementations}
\subsubsection{Library implementations}

\section{Evaluation}

\section{Project feasibility}
\subsection{Risk analysis}
\begin{table}[H]
  \centering
  \begin{tabular}{| l | l |}
    \hline
    \multicolumn{1}{|c|}{Implementation} & \multicolumn{1}{c|}{Developed by} \\
    \hline
    Eff & Bauer \& Pretnar \\ 
    \hline
    Frank & McLaughlin, Lindley and McBride \\    
    \hline 
    Koka & Leijen \\
    \hline
    Links & Hillerström and Lindley\\
    \hline
    Multicore OCaml & OCamlLabs \\
    \hline    
  \end{tabular}\caption{Languages with first-class effect handlers.}\label{tbl:efflang}
\end{table}
\subsection{Project plan}